\textbf{Proof Reconstruction in Classical Propositional Logic}

### 2017-06-16 slides/slides.tex
[ slides ] last version.
[ slides ] conjunct and atp-conjuncted fixed.

% \setbeamercolor{institute}{fg=blu}
\setbeamercolor{title}{fg=plum!80!black}
% \setbeamercolor{subtitle}{fg=blu}

@@

\usepackage[autostyle]{csquotes}
\usepackage[
    backend=biber
  , doi=false
  , eprint=false
  , isbn=false
  , natbib=true
  , sortlocale=en_US
  , style=authoryear-icomp
  , url=false
]{biblatex}
\addbibresource{ref.bib}
\renewcommand*{\nameyeardelim}{\addcomma\addspace}
\usepackage{silence}
\WarningFilter{biblatex}{Patching footnotes failed}
\WarningFilter{hyperref}{Token not allowed in a PDF string}

@@

\title[Proof Reconstruction in Classical Propositional Logic]
  {\textbf{Proof Reconstruction in Classical Propositional Logic}}
\subtitle{(work in progress)}
\date{Agda Implementors’ Meeting XXV\\
Teknikparken, Chalmers, Gothenburg, Sweden\\
May 11 2017\\
}
\author[Jonathan Prieto-Cubides]{Jonathan Prieto-Cubides\\
(joint work with Andr\'es Sicard-Ram\'irez)}
\institute{
Master in Applied Mathematics\\
Universidad EAFIT\\
Medell\'in, Colombia}
% --------------------------------------------------------------------------

@@

@@

\begin{frame}[fragile, label=athena]{\texttt{Athena}
  \footnote{\url{http://github.com/jonaprieto/athena}.}}

@@

\begin{itemize}
    \item Parsing of \texttt{TSTP} language\footnote{\label{parsing}\url{https://github.com/agomezl/tstp2agda}.}\textsuperscript{,}\footnote{\url{https://github.com/jonaprieto/tstp2agda}.}
    \item Creation\textsuperscript{\ref{parsing}} and analysis of \hyperlink{tstp-dag}{\textbf{DAG}} derivations
    \item Analysis of inference rules used in the \texttt{TSTP} derivation
    \item \texttt{Agda} code generation
\begin{table}[!ht]
\begin{center}
\scalebox{0.8}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|ll| }
\hline
\textbf{Library} & \textbf{Purpose} \\
\hline
\texttt{\hyperlink{agda-prop}{\texttt{Agda-Prop}}}& axioms and theorems of Classical Propositional Logic\\
\texttt{\hyperlink{agda-metis}{\texttt{Agda-Metis}}} & versions of the inference rules used by \texttt{Metis}\\
\hline
\end{tabular}}
}
\end{center}
\label{tab:library}
\end{table}

@@

\begin{frame}[fragile, label=agda-prop]{\texttt{Agda-Prop} Library
  \footnote{\url{https://github.com/jonaprieto/agda-prop}.}}
\begin{itemize}
  \item A data type for theorems
  \vspace*{1mm}
\begin{minted}{cagda}
data _⊢_ : (Γ : Ctxt)(φ : Prop) → Set
\end{minted}

@@

\begin{frame}[fragile, label=agda-metis]{\texttt{Agda-Metis} Library
  \footnote{\url{https://github.com/jonaprieto/agda-metis}.}}
\vfill
\begin{table}[!ht]
\begin{center}
\scalebox{0.7}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|lll|}
\hline
\textbf{Rule} & \textbf{Purpose} &\textbf{Theorem}\\
\hline
\texttt{canonicalize} & transforms formulas to CNF, DNF or NNF &\texttt{atp-canonicalize}\\
\texttt{clausify} & performs clausification &\texttt{atp-clausify}\\
\hyperlink{atp-conjunct}{\texttt{conjunct}} & takes a formula from a conjunction &\hyperlink{atp-conjunct}{\texttt{atp-conjunct}}\\
\texttt{negate} & applies negation to the formula &\texttt{atp-negate}\\
\texttt{resolve} & applies theorems of resolution &\texttt{atp-resolve}\\
\texttt{simplify} & applies over a list of formula to simplify them &\texttt{atp-simplify}\\
\texttt{strip} & splits a formula into subgoals &\texttt{atp-strip}\\[1ex]
\hline
\end{tabular}}
}
\end{center}
\label{tab:agda-metis-table}
\end{table}
\vfill
\end{frame}

\begin{frame}[fragile, label=atp-conjunct]{\texttt{Agda-Metis}: Conjunct Inference
  \footnote{\url{https://github.com/jonaprieto/agda-metis}.}}
\begin{itemize}
\item Definition
  \begin{equation*}
  \text{conjunct}(ϕ₁ ∧  ϕ₂ ∧ ⋯ ∧ \mathbf{ϕᵢ} ∧ ⋯ ∧ \phi_n, \mathbf{ϕᵢ}) ⟶ \mathbf{ϕᵢ}
  \end{equation*}

\item Function
\vspace*{1mm}
\begin{minted}{cagda}
conjunct : Prop → Prop → Prop
conjunct (φ ∧ ψ) ω with ⌊ eq φ ω ⌋ | ⌊ eq ψ ω ⌋
... | true  | _     = φ
... | false | true  = ψ
... | false | false = conjunct φ ω
conjunct φ ω        = φ
\end{minted}
\item Theorem
\vspace*{1mm}
\begin{minted}{cagda}
atp-conjunct
  : ∀ {Γ} {φ}
  → (ω : Prop)
  → Γ ⊢ φ
  → Γ ⊢ conjunct φ ω
\end{minted}
\end{itemize}
\end{frame}

\begin{frame}[fragile, label=atp-conjunct-proof]{A proof of \texttt{atp-conjunct} theorem}
\vfill
\begin{minted}{cagda}
atp-conjunct
  : ∀ {Γ} {φ}
  → (ω : Prop)
  → Γ ⊢ φ
  → Γ ⊢ conjunct φ ω

atp-conjunct {Γ} {φ ∧ ψ} ω Γ⊢φ
  with ⌊ eq φ ω ⌋ | ⌊ eq ψ ω ⌋
... | true  | _     = ∧-proj₁ Γ⊢φ
... | false | true  = ∧-proj₂ Γ⊢φ
... | false | false =
  atp-conjunct {Γ = Γ} {φ = φ} ω (∧-proj₁ Γ⊢φ)
atp-conjunct {_} {Var x}  _ = id
atp-conjunct {_} {⊤}      _ = id
atp-conjunct {_} {⊥}      _ = id
atp-conjunct {_} {φ ∨ ψ}  _ = id
atp-conjunct {_} {φ ⇒ ψ}  _ = id
atp-conjunct {_} {φ ⇔ ψ}  _ = id
atp-conjunct {_} {¬ φ}    _ = id
\end{minted}
\vfill
\gobackreconstruction
\end{frame}

@@

\begin{itemize}
  \item The problem is $p ∧ q ⊢ q ∧ p$
  \item In \texttt{TPTP} format
\begin{tptp}
fof(a, axiom, p & q).
fof(goal, conjecture, q & p).
\end{tptp}
  \item A natural deduction proof
\begin{columns}
\begin{column}{0.8\textwidth}
 \begin{prooftree}
  \AxiomC{$ϕ ∧ ψ$}
  \RightLabel{$∧$-\texttt{proj}$₁$}
  \UnaryInfC{$ϕ$}

  \AxiomC{$ϕ ∧ ψ$}
  \RightLabel{$∧$-\texttt{proj}$₂$}
  \UnaryInfC{$ψ$}
  \RightLabel{$∧$-\texttt{intro}}
  \BinaryInfC{$ψ ∧ ϕ$}
  \end{prooftree}
\end{column}
\begin{column}{0.2\textwidth}  %%<--- here
   \begin{tikzpicture}[overlay, remember picture, scale=2]
% \tikzset{shift={(current page.north)}, xshift=cm,yshift=-2cm}
\node[vertex, fill=black] (d1) {};
\node[vertex, fill=black, right=0.5cm of d1 ] (d2) {};
\node[vertex, fill=black, below=0.2cm of d1] (d3) {};
\node[vertex, fill=black, below=0.2cm of d2] (d4) {};
\node[vertex, fill=black, below right= 0.2cm and  0.25cm of d3] (d5) {};
\draw[-, thick] (d1) to (d3);
\draw[-, thick] (d2) to (d4);
\draw[-, thick] (d3) to (d5);
\draw[-, thick] (d4) to (d5);
\end{tikzpicture}
\end{column}
\end{columns}

@@

\begin{frame}[label=contributions]{Contributions}
\vfill
\begin{table}[!ht]
\begin{center}
\scalebox{0.8}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|ll|}
\hline
\textbf{Name} & \textbf{Purpose} \\
\hline
\href{http://github.com/jonaprieto/agda-metis}{\texttt{Agda-Metis}}   & Implementation of \texttt{Metis} inference rules\\
\href{http://github.com/jonaprieto/agda-prop}{\texttt{Agda-Prop}}     & Syntax and theorems of Classical Propositional Logic\\
\href{http://github.com/jonaprieto/athena}{\texttt{Athena}}           & Translator for \texttt{Metis} \texttt{TSTP} files to \texttt{Agda}\\
\href{http://github.com/jonaprieto/online-atps}{\texttt{OnlineATPs}}  & Client to use ATPs from \texttt{SystemOnTPTP} of \texttt{TPTP} World\\
\href{http://github.com/jonaprieto/prop-pack}{\texttt{Prop-Pack}}     & Collection of \texttt{TPTP} problems to test \texttt{Athena}\\[1ex]
\hline
\end{tabular}}
}
\end{center}
\label{tab:library}
\end{table}
\vfill
\end{frame}

@@

\begin{frame}[fragile, label=references, allowframebreaks]{References}
\printbibliography
\end{frame}

>>>>

% \setbeamercolor{institute}{fg=blu}
\setbeamercolor{title}{fg=blu}
% \setbeamercolor{subtitle}{fg=blu}

@@

\usepackage[autostyle]{csquotes}
\usepackage[
    backend=biber
  , doi=false
  , eprint=false
  , isbn=false
  , natbib=true
  , sortlocale=en_US
  , style=authoryear-icomp
  , url=true
]{biblatex}
\addbibresource{ref.bib}
\renewcommand*{\nameyeardelim}{\addcomma\addspace}
\usepackage{silence}
\WarningFilter{biblatex}{Patching footnotes failed}
\WarningFilter{hyperref}{Token not allowed in a PDF string}

@@

\title[Proof Reconstruction in Classical Propositional Logic]
  {\textbf{Proof Reconstruction in Classical Propositional Logic}}
\subtitle{(work in progress)}
\date{May 11, 2017
}
\author[Jonathan Prieto-Cubides]{Jonathan Prieto-Cubides\\
(joint work with Andr\'es Sicard-Ram\'irez)}
\institute{
Master in Applied Mathematics\\
Universidad EAFIT\\
Medell\'in, Colombia}
% --------------------------------------------------------------------------

@@

% Now each step paused.

@@

\begin{frame}[fragile, label=athena]{\texttt{Athena}~\citep{Athena}}

@@

\begin{itemize}
    \item Parsing of \texttt{TSTP} language\footnote{\label{parsing}\url{https://github.com/agomezl/tstp2agda}.}
    \item Creation\textsuperscript{\ref{parsing}} and analysis of \hyperlink{tstp-dag}{\textbf{DAG}} derivations
    \item Analysis of inference rules used in the \texttt{TSTP} derivation
    \item \texttt{Agda} code generation
\begin{table}[!ht]
\begin{center}
\scalebox{0.8}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|ll| }
\hline
\textbf{Library} & \textbf{Purpose} \\
\hline
\texttt{\hyperlink{agda-prop}{\texttt{Agda-Prop}}}& axioms and theorems of Classical Propositional Logic\\
\texttt{\hyperlink{agda-metis}{\texttt{Agda-Metis}}} & versions of the inference rules used by \texttt{Metis}\\
\hline
\end{tabular}}
}
\end{center}
\label{tab:library}
\end{table}

@@

\begin{frame}[fragile, label=agda-prop]{\texttt{Agda-Prop} Library~\citep{AgdaProp}}
\begin{itemize}
  \item A data type for theorems
  \vspace*{1mm}
\begin{minted}{cagda}
data _⊢_ : (Γ : Ctxt)(φ : Prop) → Set
\end{minted}

@@

\begin{frame}[fragile, label=agda-metis]{\texttt{Agda-Metis} Library
  ~\citep{AgdaMetis}}
\vfill
\begin{table}[!ht]
\begin{center}
\scalebox{0.7}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|lll|}
\hline
\textbf{Rule} & \textbf{Purpose} &\textbf{Theorem}\\
\hline
\texttt{canonicalize} & transforms formulas to CNF, DNF or NNF &\texttt{atp-canonicalize}\\
\texttt{clausify} & performs clausification &\texttt{atp-clausify}\\
\hyperlink{atp-conjunct}{\texttt{conjunct}} & takes a formula from a conjunction &\hyperlink{atp-conjunct}{\texttt{atp-conjunct}}\\
\texttt{negate} & append negation symbol to the formula &\texttt{atp-negate}\\
\texttt{resolve} & applies theorems of resolution &\texttt{atp-resolve}\\
\texttt{simplify} & applies over a list of formula to simplify them &\texttt{atp-simplify}\\
\texttt{strip} & splits a formula into subgoals &\texttt{atp-strip}\\[1ex]
\hline
\end{tabular}}
}
\end{center}
\label{tab:agda-metis-table}
\end{table}
\vfill
\end{frame}

\begin{frame}[fragile, label=atp-conjunct]{\texttt{Agda-Metis}: Conjunct Inference
  \footnote{\url{https://github.com/jonaprieto/agda-metis}.}
  }

\vfill
\begin{itemize}
\item Definition
\begin{equation*}
\text{\color{plum!40!black}conjunct}(ϕ₁ ∧ ⋯ ∧ \phi_n, \omega)
= \begin{cases}
\phi_{i} \ \ \ \ \text{if }\omega\text{ is equal to some }\phi_i\\
\phi    \ \ \ \ \text{otherwise}
\end{cases}
\end{equation*}

\pause
\item Inference rules involved
\begin{center}
\begin{columns}
\begin{column}{0.45\textwidth}

 \begin{prooftree}
  \AxiomC{$ϕ ∧ ψ$}
  \RightLabel{\scriptsize $∧$-\texttt{proj}$₁$}
  \UnaryInfC{$ϕ$}
  \end{prooftree}
\end{column}

\begin{column}{0.45\textwidth}
 \begin{prooftree}
  \AxiomC{$ϕ ∧ ψ$}
  \RightLabel{\scriptsize $∧$-\texttt{proj}$₂$}
  \UnaryInfC{$ψ$}
  \end{prooftree}
\end{column}

\begin{column}{0.1\textwidth}  %%<--- here
   \begin{tikzpicture}[overlay, remember picture, scale=2]
% \tikzset{shift={(current page.north)}, xshift=cm,yshift=-2cm}
\node[vertex, fill=black] (d1) {};
\node[vertex, fill=black, below left=0.2cm and 0.1cm of d1] (d2) {};
\node[vertex, fill=black, below right=0.2cm and 0.1cm of d1] (d3) {};

\node[vertex, fill=black, below left=0.2cm and 0.1cm of d3] (q) {};
\node[vertex, fill=black, below right=0.2cm and 0.1cm of d3] (w) {};

\node[vertex, fill=black, below right=0.2cm and 0.1cm of q] (s) {};

\draw[-, thick] (d1) to (d2);
\draw[-, thick] (d1) to (d3);
\draw[-, thick] (d3) to (q);
\draw[-, thick] (d3) to (w);
\draw[-, thick] (q) to (s);
\end{tikzpicture}
\end{column}
\end{columns}
\end{center}
\vskip 2mm

\item Target example
\[ φ := (¬ p ∧ (¬ q ⇔ ¬ r)) ∧ (p ∧ (¬ q ⇔ ¬r))\]

\begin{itemize}

\item
\begin{minted}{cagda}
 conjunct(φ , ¬ q) ≡ ϕ
\end{minted}
\vskip 3mm
\item
\begin{minted}{cagda}
 conjunct(φ , ¬ q ⇔ ¬ r) ≡ ¬ q ⇔ ¬ r
\end{minted}
\vskip 3mm

\item
\begin{minted}{cagda}
 conjunct(φ , p)   ≡ p
\end{minted}

\end{itemize}

\begin{frame}[fragile, label=atp-conjunct]

\begin{minted}{cagda}
data ConjView : Prop → Set where
  conj  : (φ₁ φ₂ : Prop) → ConjView (φ₁ ∧ φ₂)
  other : (φ : Prop)     → ConjView φ

conj-view : (φ : Prop) → ConjView φ
conj-view (φ ∧ ψ) = conj _ _
conj-view φ       = other _

data Step  : Set where
  pick  : Step
  proj₁ : Step
  proj₂ : Step

Path : Set
Path = List Step

conjunct-path : (φ : Prop) → (ω : Prop) → Path → Path
conjunct-path φ ω path with ⌊ eq φ ω ⌋
... | true  = path ∷ʳ pick
... | false with conj-view φ
...   | other _   = []
...   | conj φ₁ φ₂ with conjunct-path φ₁ ω []
...     | subpath@(_ ∷ _) = (path ∷ʳ proj₁) ++ subpath
...     | [] with conjunct-path φ₂ ω []
...         | subpath@(_ ∷ _) = (path ∷ʳ proj₂) ++ subpath
...         | []              = []
\end{minted}

\begin{frame}[fragile, label=atp-conjunct-proof]{The \texttt{conjunct} function and its theorem, \texttt{atp-conjunct}}
\vfill
\begin{minted}{cagda}
conjunct : (φ : Prop) → (ω : Prop) → Prop
conjunct φ ω with conj-view φ | conjunct-path φ ω []
... | _          | []           = φ
... | conj φ₁ φ₂ | pick  ∷ path = φ
... | conj φ₁ φ₂ | proj₁ ∷ path = conjunct φ₁ ω
... | conj φ₁ φ₂ | proj₂ ∷ path = conjunct φ₂ ω
... | other .φ   | _            = φ

atp-conjunct
  : ∀ {Γ} {φ}
  → (ω : Prop)
  → Γ ⊢ φ
  → Γ ⊢ conjunct φ ω

atp-conjunct {Γ} {φ} ω Γ⊢φ
  with conj-view φ | conjunct-path φ ω []
...  | _          | []           = Γ⊢φ
...  | conj φ₁ φ₂ | pick  ∷ path = Γ⊢φ
...  | conj φ₁ φ₂ | proj₁ ∷ path = atp-conjunct ω (∧-proj₁ Γ⊢φ)
...  | conj φ₁ φ₂ | proj₂ ∷ path = atp-conjunct ω (∧-proj₂ Γ⊢φ)
...  | other .φ   | (_ ∷ _)      = Γ⊢φ
\end{minted}
\vfill
\gobackreconstruction
\end{frame}

@@

\begin{itemize}
  \item The problem is $p ∧ q ⊢ q ∧ p$
  \item In \texttt{TPTP} format
\begin{tptp}
$ cat problem.tptp
fof(a, axiom, p & q).
fof(goal, conjecture, q & p).
\end{tptp}
  \item How to use Athena with your problem
  \vspace*{5mm}
\begin{tikzpicture}
\node[rectangle, text=white,fill=black, align=left, font=\small]
{
\texttt{\$ metis --show proof problem.tptp > problem.tstp}\hspace{1cm}\\
\texttt{\$ athena problem.tstp}\\
\texttt{\$ agda problem.agda}
  };
\end{tikzpicture}

@@

\begin{frame}[label=contributions]{Contributions}
\vfill
\begin{table}[!ht]
\begin{center}
\scalebox{0.8}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|ll|}
\hline
\textbf{Name} & \textbf{References} \\
\hline
\href{http://github.com/jonaprieto/agda-metis}{\texttt{Agda-Metis}}   & \citep{AgdaMetis}\\
\href{http://github.com/jonaprieto/agda-prop}{\texttt{Agda-Prop}}     & \citep{AgdaProp}\\
\href{http://github.com/jonaprieto/athena}{\texttt{Athena}}           & \citep{Athena}\\
\href{http://github.com/jonaprieto/online-atps}{\texttt{OnlineATPs}}  & \citep{OnlineATPs}\\
\href{http://github.com/jonaprieto/prop-pack}{\texttt{Prop-Pack}}     & \citep{ProPack}\\[1ex]
\hline
\end{tabular}}
}
\end{center}
\label{tab:library}
\end{table}
\vfill
\end{frame}

@@

\begin{frame}[allowframebreaks]{References}
\printbibliography
\end{frame}
